Accelerating Burn Speed of Birch Wood Popsicle Sticks

### 2022-05-01 paper/paper.Rmd
Update paper and change Skillings Mack MC procedure to use Bonferroni adjustment

---
title: "Accelerating Burn Speed of Birch Wood Popsicle Sticks"
author: "E. Croffoot-Suede, S. Kim"
date: "4/25/2022"
output:
  pdf_document:
    pandoc_args: --listings
    includes:
      in_header: preamble.tex
---

@@

@@

PUT SOME HYPOTHESES HERE!

@@

@@

To ensure that we could find statistically significant results, we ran a power calculation. We assumed that we would use six time blocks, and we chose an alpha level for this of $\alpha = 0.05$, which is the probability of a Type I error (false rejection). We chose a Type II error probability of 0.1. Based on this, we calculated $N = 24$, meaning that we should use 6 blocks of 4 observations each. MAKE SURE TO THROW IN MORE INFO ON TAKING THE BIGGER N HERE!

@@

@@

```{R normality, fig.show="hold", out.width="50%", fig.height = 5, echo = FALSE}
plot(aov(distance ~ type * run, data = fire), which = 1, main = "Figure 1: Residuals vs Fitted")
plot(aov(distance ~ type * run, data = fire), which = 2, main = "Figure 2: Normal Q-Q")
```

In Figure 1, it is clear that there is a serious issue with the equal variance condition. There is a distinct "funnel" shape from left to right, where the variance at the fitted values around 80 to 90 are much larger than the variance at 105 to 110.

In Figure 2, there are flaired tails that indicate an overdispersed data distribution compared to a normal distribution. That is the distribution was "squished" toward the center compared to the normal distribution.

@@

@@

We got the following comparisons table, which contains the cutoff values at an $\alpha$ level of 0.1.^[We use the \nth{5} and \nth{95} percentiles because these are the $\alpha/2$ levels.] This is essentially a method to calculate the distance from the median and tell whether there is a significant difference between two groups.

| Comparison        | Actual | Medians of Simulated U’s | \nth{5} Percentile | \nth{95} Percentile |
|-------------------|--------|--------------------------|----------------|-----------------|
| Ethanol-Dry       | 2.5    | 18.5                     | 8.0            | 28.5            |
| Ethanol-Acetone   | 2.0    | 18.0                     | 7.5            | 28.5            |
| Ethanol-Isopropyl | 35.0   | 18.0                     | 8.0            | 27.5            |
| Acetone-Dry       | 17.0   | 17.5                     | 7.0            | 28.5            |
| Acetone-Isopropyl | 19.0   | 18.0                     | 7.0            | 28.5            |
| Isopropyl-Dry     | 23.0   | 18.0                     | 7.0            | 28.0            |

@@

```{r boxplot, echo = FALSE}
boxplot(distance ~ type, data = fire, ylab="Distance (mm)", 
        xlab="Accelerant", main="Figure 3: Unburnt Distance by Accelerant")
```

@@

@@

Our goal is to maximize the amount burnt, so the metric for success is to minimize the unburnt distance. We saw in the boxplot from Figure 3 that the best way to decrease unburnt distance is to use acetone. Table 1 tells us that there was no significant difference between acetone, dry, or isopropyl alcohol. Therefore, we recommend that to maximize the amount burnt, to soak birch popsicle sticks in acetone, isoproply alcohol, or no accelerant at all.

>>>>

---
title: "Accelerating Burn Speed of Birch Wood Popsicle Sticks"
author: "E. Croffoot-Suede, S. Kim"
date: "4/25/2022"
output:
  pdf_document:
    pandoc_args: --listings
    includes:
      before_body: title.sty
      in_header: preamble.tex
---

@@

We predicted that using any liquid accelerant would be faster than no liquid accelerant, and that in comparing individual treatments, there would be no significant difference between the accelerants, but there would be a significant difference between no accelerant and with accelerant.

@@

@@

This gives us a $4!$ experimental design, since we have four treatment effects and no other factors.

@@

To ensure that we could find statistically significant results, we ran a power calculation. We assumed that we would use six time blocks, and we chose an alpha level for this of $\alpha = 0.05$, which is the probability of a Type I error (false rejection). We chose a Type II error probability of 0.1. Given a $k = 4$ (given four treatment), the calculated sample size was 20, but given a $k = 6$ (given six time blocks), our calculated sample size was 24. Given our two-way layout, we should take the larger $N$ calculated out of $k = 4, 6$. Based on this, we used $N = 24$, meaning that we should use 6 blocks of 4 observations each.

To generate the order in which we burnt our sticks, we used Python to generate an empty dataframe that had the treatments for 24 runs. We then randomly assigned each of the rows to one of six blocks. This gives us an experimental design with six blocks with four treatments per block.

This method of block generation, however, means that it is possible to have multiple iterations of one treatment and none of other treatments in the same block. For example, it is possible to have "acetone", "acetone", "isopropyl", and "dry" in a single block, meaning that "ethanol" would not be tested in this block. This gives us arbitrary unbalanced data.

@@

We began by examining the data. To do so, we generated a boxplot of distances, as well as a boxplot of treatment against distance.

```{r eda, fig.show="hold", out.width="50%", fig.height = 5, echo = FALSE}
boxplot(fire$distance, ylab="Distance (mm)", main = "Figure 1: Unburnt distance")
boxplot(distance ~ type, data = fire, ylab="Distance (mm)", 
        xlab="Accelerant", main = "Figure 2: Unburnt distance by accelerant")
```

Figure 1 shows the range of response values, but to get a better view of the treatment effects, we also made Figure 2, a plot of unburnt distance by accelerant. We are not using the blocking factor as a factor of interest, so the only plot to be made using a factor against response is with treatment, as shown in Figure 2.

We begin with the model

$$
y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij}
$$

with $i = 1, 2, 3, 4$ and $j = 1, 2, 3, 4, 5, 6$, assuming that the conditions for an ANOVA are met.

@@

```{R normality, fig.show="hold", out.width="50%", fig.height = 5, echo = FALSE}
plot(aov(distance ~ type * run, data = fire), which = 1, main = "Figure 3: Residuals vs Fitted")
plot(aov(distance ~ type * run, data = fire), which = 2, main = "Figure 4: Normal Q-Q")
```

In Figure 3, it is clear that there is a serious issue with the equal variance condition. There is a distinct "funnel" shape from left to right, where the variance at the fitted values around 80 to 90 are much larger than the variance at 105 to 110.

In Figure 4, there are flaired tails that indicate an overdispersed data distribution compared to a normal distribution. That is the distribution was "squished" toward the center compared to the normal distribution.

@@

Thusly, our model is updated to

$$
y_{ij} = \theta + \alpha_i + \beta_j + \epsilon_{ij}
$$

with $i = 1, 2, 3, 4$ and $j = 1, 2, 3, 4, 5, 6$, to accommodate this new procedure, given that the conditions for the ANOVA are not met.

The Skillings-Mack procedure requires that the $N$ random variables are mutually independent. That is, $X_{ij1}, ..., X_{ijc_{ij}}$, $i = 1, ..., n$, and $j = 1, ..., k$ are mutually independent. We believe this to be true for our data given the collection method. It also requires that for each fixed $(i, j)$, the $c_{ij}$ random variables are a random sample from a continuous distribution function $F_{ij}$. We also believe this to be true because we used a random sample out of the population of popsicle sticks in the world. Lastly, the treatment response should be a continuous variable such that "less than", "equal to", and "greater than" can be placed into context. We believe that "length" in millimeters as a response variable satisfies this condition.

The hypotheses for the Skillings-Mack procedure are as follows:

$$
H_0: [\alpha_1 = \alpha_2 = \alpha_3 = \alpha_4],
$$

$$
H_A: [\alpha_1, \alpha_2, \alpha_3, \alpha_4 \text{ not all equal}],
$$

when $c_{ij} \equiv 1$, for $i = 1, 2, 3, 4$ and $j = 1, 2, 3, 4, 5, 6$.

@@

We got the following comparisons table, which contains the cutoff values at an $\alpha$ level of 0.1.^[We use the $\frac{0.1}{6}$ and the $1 - \frac{0.1}{6}$, since these are the Bonferroni adjustments given six tests. Not using a Bonferroni adjustment makes the cutoff values for the 90% confidence interval simply $\frac{\alpha}{2}$ and $1 - \frac{\alpha}{2}$.] This is essentially a method to calculate the distance from the median and tell whether there is a significant difference between two groups.

| Comparison        | Actual | Medians of Simulated U’s | $\frac{0.1}{6}$ Percentile | $1 - \frac{0.1}{6}$ Percentile |
|-------------------|--------|--------------------------|----------------|-----------------|
| Ethanol-Dry       | 2.5    | 18.5                     | 4.833333       | 30.66667        |
| Ethanol-Acetone   | 2.0    | 18.0                     | 5.000000       | 30.50000        |
| Ethanol-Isopropyl | 35.0   | 18.0                     | 5.500000       | 30.50000        |
| Acetone-Dry       | 17.0   | 17.5                     | 4.333333       | 31.16667        |
| Acetone-Isopropyl | 19.0   | 18.0                     | 4.333333       | 31.16667        |
| Isopropyl-Dry     | 23.0   | 18.0                     | 4.000000       | 30.00000        |

@@

```{r boxplot, echo = FALSE}
boxplot(distance ~ type, data = fire, ylab="Distance (mm)", 
        xlab="Accelerant", main="Figure 5: Unburnt Distance by Accelerant")
```

@@

In retrospect, we should have designed our experiment to use a randomized complete block design, instead of an arbitrary incomplete block design. In a randomized complete block design, we would still have six blocks and four treatments, but each block would contain one observation for each of the possible treatments. This would have made our analysis easier. If our data had been collected as a randomized complete block design, we would have used the Friedman procedure for differences in medians using a randomized complete block design, and assuming a significant effect, we would have followed it up with the distribution-free, two-sided, all-treatments Wilcoxon, Nemenyi, McDonald-Thompson multiple comparisons procedure based on Friedman rank sums in a general configuration.

However, we believe that our pairwise multiple comparisons procedure would still be valid given data obtained in an arbitrary incomplete block design. For example, suppose we were looking at data for a factory that produces products in batches, but which does not produce every batch every day. This would generate data in an arbitrary incomplete block design, where a metric of the product is the response, the batch is the treatment, and the day is the block. Using our procedure, we believe that we could diminish the block effect of days in the analysis and come up with an appropriate pairwise multiple comparison between the batches.

We believe that the results can be generalized to a wider population to what accelerants are the most dangerous for increasing flammability in traditionally flammable materials, like wood.

@@

Our goal is to maximize the amount burnt, so the metric for success is to minimize the unburnt distance. We saw in the boxplot from Figure 5 that the best way to decrease unburnt distance is to use acetone. Table 1 tells us that there was no significant difference between acetone, dry, or isopropyl alcohol. Therefore, we recommend that to maximize the amount burnt, to soak birch popsicle sticks in acetone, isopropyl alcohol, or no accelerant at all.

Based on this, we reject our original hypothesis that using any accelerant is better than no accelerant in favor of the theory that 75.5% ethanol alcohol has a negative effect on burning distance.
